Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-27 rebuttal.tex

\textbf{(6) There is only limited statistical assessment/validation of the models. Statements made in the
text that R2 values are “reasonable” and the models perform “relatively well” are difficult to
reconcile. Rather, Table 2 indicates that the R2 values were rather poor in several instances.}

The text has been updated to read more quantitatively and avoid ambiguous statements

\textbf{(7) It appears that there are systematic prediction errors for pairs. At low value ranges (that are
not relevant for activity cliff assessment), SALI values are consistently over-predicted. By
contrast, at high value ranges (that are relevant), SALI values are consistently under-predicted, in
part significantly. This is not investigated, although the apparent errors at high value ranges are
the perhaps most critical features of the reported RF models.}

Yes, there do appear to be such systematic errors - at this stage it's not clear what is the
cause. However we have noted that while it is clearly evident for the Cavalli dataset, it is not
as obvious for the ChEMBL datasets and in some cases is the reverse of what we see in the Cavalli
case. For example, for the Dai dataset, we observed that lower SALI values are overestimated
(whereas the Cavalli predictions are underestimaetd at low SALI values).

We hypothesize that the descriptors employed do not differentiate between observations with low and
high SALI values very well, leading to overestimation of small values and underestimation of high
SALI values. We have justified this by comparing descriptor distance distributions between the
group of observations with low SALI values and those with high SALI values. The two distributions
nearly completely overlap, supporting this hypothesis.

@@

\textbf{(8) In the section “Extending a landscape”, predictions are finally reported for only three (!)
molecules taken out of each data set. Carefully put, the predictions are heterogeneous. On the basis
of the results reported for this little bit of evaluation, one would be hard pressed to make a case
for the ability of the approach to predict activity cliffs with any certainty.}

We have updated this section to include several more hold out molecules. It should be noted that this section
examines the model from a different viewpoint compared to Section 4. Specifically, Section 4.1
develops models in which we remove molecules from the original dataset. Given the size of the
original datasets, removing more than 10 or so, leads smaller pairwise datasets. Simultaneously, the
predictions for this validation stratgey are not just for the 10 hold out molecules - it will be 10
$\times m$ where $m$ is the number of the molecules in the original datasets. Thus the number of
predictiosn range from 250 to 340 values. Summarizing more than a few of these hold out molecules
becomes cumbersome.

@@

Certainly, with respect to the previous comment, a pair of molecules with $T_c = 0.29$ would likely
not be regarded as a cliff - I specifically note that such a case is not a significant cliff. To
address this aspect of SALI values, I have presented plots of dissimilarity versus activity ratio
(Figures 7 and 9) which allow one to more directly determine ``significant'' versus
``insignificant'' cliffs.

\textbf{Furthermore, there might be many instances of activity and fingerprint similarity
  relationships between pairs of compounds that might yield high pair-based SALI scores dominated by
  either the activity or similarity component, without forming a “true” activity cliff. The
  conclusion of this reviewer is that it is currently not possible to capture activity cliffs with
  any degree of certainty on the basis of the pair calculations, as reported.}

While it is true that the SALI definition itself allows for high SALI values to be generated for
pairs of molecules that would be small/insignificant cliffs (highly similar structures with small
difference in activity), the methodology does let one rank pairs of molecules and as indicated in
Figures 7 and 9, one can compare the strucural similarity and activity difference (or ratio) in 2D,
allowing one to separate ``true'' cliffs from apparent ones.

In addition, Figures 8 and 10 do suggest that the approach does capture actual activity cliffs - it
so happens that the datasets considered do not contain too many of them, to allow one to provide a
large scale validation.

As noted in Section 5, predictive performance can likely be increased by a larger, more diverse
descriptor pools as well as alternative modeling methods. The goal of this paper is to suggest that
the proposed pairwise methodology does let as address the identifcation of activity cliffs in a
prospective manner.

@@

\textbf{4.  SALI values needs to be determined for pairs of molecules - is that not the essence of
  the methodology?}

Yes. For a given dataset, with measured activity values, one can indeed evaluate the SALI
values. The proposed methodology takes these ``measured'' SALI values for a training set and then
predicts the SALI values for a new molecule with each member of the training set. As with
traditional QSAR models, a complete validation would require one to measure the activity of the new
molecule and evaluate the actual SALI values. In absence of experimental activities for the test
set, we resort to a training/test split, which is an accepted approach in QSAR modeling.

\textbf{5.  There is no discussion/direction of how to use the methodology in new applications.}

Section 4.1 in the original manuscript specifically addresses how one might use the proposed
methodology. This has been made more explicit in the updated manuscript.

### 2012-07-27 paper.tex
final set of edits for resubmission

Note that the predictions on the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity cliffs, which are likely not
very interesting. Given that observation, it is encouraging to observe that all three models perform
relatively well at the higher end of the SALI spectrum. For this example, the three most significant
activity cliffs are in fact not very significant activity cliffs in an absolute sense - the $T_c$
for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
5.11 log units respectively.

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted (from the training set) SALI
values using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{absdiff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, the most significant cliffs
exhibit significant variance. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$
as the aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and
Figure \ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported
activity data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of
these models employed the entire dataset (rather than splitting into a training and testing set) and
exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
and $R^2$ (0.18 to 0.34).. This can be partially ascribed to the small sizes of the dataset. It
should be note that this methodology is not aimed at modeling the original, usually small, SAR
dataset. It is well known (and evident from the current analyses) that small datasets do not
necessarily lead to very robust or reliable machine learning models. Instead, the methodology
focuses on the \emph{pairwise} dataset -- which are usually large enough for a statistically
significant model.

@@

The results described above suggest that predicting pairwise SALI values is a useful way to identify
whether a new molecule will form an activity cliff with one or more members of the training set. As
with all QSAR models, one is limited by the nature of the underlying data, both in terms of accuracy
as well as applicability. We have selected a few ChEMBL assays for the purposes of
highlighting this approach, we have found that for some assays, the model built on pairwise data
performs worse than the model built on the original data. Given that most of the datasets are small
in size (less than 40 molecules), the poor performance of the initial model is not always
surprising. However, in cases where the pairwise model does not exhibit improved performance, we
believe that predictive modeling of activity cliffs in the original dataset will not be fruitful.

@@

@@

\ctable[cap={}, caption={Distribution of log(SALI) values for the
  three ChEMBL datasets.}, figure, botcap,
label={fig:logsalidist}]    
{c}
{}  
{\includegraphics[width=\linewidth]{logsalidist}}

>>>>

Note that the predictions on the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity cliffs, which are likely not
very interesting. Given that observation, it is encouraging to observe that all three models perform
relatively well at the higher end of the SALI spectrum. For this example, the three most significant
activity cliffs are in fact not very significant activity cliffs in an absolute sense - the $T_c$
for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
5.11 log units respectively.  But more importantly, it is clear from all three panels that the there
appears to be two different systematic prediction errors - smaller SALI values are overestimated
whereas larger values are underestimated.  A possible explanation for this behavior is that the
descriptors do not sufficiently differentiate observations with low and high SALI values. In other
words, the descriptor profile for observations with low SALI values is sufficiently similar to those
with high SALI values. To test this out, we split the Cavalli dataset into two groups - low and high
- using the median SALI value as the cutoff. We then evaluated the pairwise descriptor distance
distributions for each each of the groups, using the Euclidean distance between all descriptors in
the pool. Figure \ref{fig:cavalli-desc-dist} indicates that the distance distributions are
essentially identical. Though random forest models employ a subset of the descriptor pool for each
tree, the same reasoning can be applied to these subsets. These trends appear in the other datasets
(Figure \ref{fig:rfmodel-chembl}) but to varying degrees and not always consistently. For example,
in the Costanzo dataset, we see that the higher SALI values are somewhat underestimated. The concept
of descriptor similarity could also be applied to these cases. But a more fundamental explanation of
this behavior is not available at this time.

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values (for the training
set) using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{diff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, systematic prediction
errors are apparent. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$ as the
aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and Figure
\ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported activity
data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of these
models employed the entire dataset (rather than splitting into a training and testing set) and
exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
and $R^2$ (0.18 to 0.34). This can be partially ascribed to the small sizes of the dataset. It
should be note that this methodology is not aimed at modeling the original, usually small, SAR
dataset. It is well known (and evident from the current analyses) that small datasets do not
necessarily lead to very robust or reliable machine learning models. Instead, the methodology
focuses on \emph{pairwise} datasets -- which are usually large enough for a statistically
significant model.

@@

The results described above suggest that predicting pairwise SALI values is a useful way to identify
whether a new molecule will form an activity cliff with one or more members of the training set. As
with all QSAR models, one is limited by the nature of the underlying data, both in terms of accuracy
as well as applicability. We have selected a few ChEMBL assays for the purposes of highlighting this
approach. Given that most of the datasets are small in size (30 to 50 molecules), models built on
the actual activity data do not perform very well. In contrast the models built on the pairwise data
usually exhibit improved performance - which could certainly be ascribed to the larger dataset
involved. However, in cases where the pairwise model does
not exhibit improved performance, we believe that predictive modeling of activity cliffs in the
original dataset will not be fruitful.

@@

\ctable[cap={}, caption={A comparison of the pairwise descriptor distribution for low and high SALI
  values in the Cavalli dataset. This figure employed the descriptor values generated using the
  $f_{diff}$ aggregation function and Euclidean distances were evaluataed using all the descriptors
  in the pool. The \emph{low} group is defined as those observations with SALI
  less than 2.03 and the remainder is assigned to the \emph{high} group.},
label={fig:cavalli-desc-dist}, botcap, figure]
{c}
{}
{
\includegraphics[width=\linewidth]{cavalli-desc-dist}
}

@@

\ctable[cap={}, caption={Distribution of observed log(SALI) values for the
  three ChEMBL datasets.}, figure, botcap,
label={fig:logsalidist}]    
{c}
{}  
{\includegraphics[width=\linewidth]{logsalidist}}
